Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-05 model.tex
adding some figures and editing

\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA){\color{red} NOT EDITABLE}}\label{sec:lma}

When a generating process exhibits deterministic predictive nonlinear structure a common approach is to reconstruct the generating processes dynamics using a technique called delay-coordinate embedding and then use the geometric structure of this reconstruction as a proxy for the information transmission process of the underlying process to predict the future. In this section we discuss the theory and implementation of this forecasting scheme.

 \subsubsection{Reconstructing hidden dynamics {\color{red} NOT EDITABLE}}

Delay-coordinate embedding allows one to reconstruct a system's full
state-space dynamics from a \emph{single} scalar time-series
measurement---provided that some conditions hold regarding that data.
Specifically, if the underlying dynamics and the measurement
function---the mapping from the unknown state vector $\vec{X}$ to the
scalar value $x$ that one is measuring---are both smooth and generic,
Takens~\cite{takens} formally proves that the delay-coordinate map
\[
F(\tau,m)(X) = ([X_{i+\tau,obs} ~ X_{i+\tau,obs} ~ \dots ~X_{i+m\tau,obs}])
\]
from a $d$-dimensional smooth compact manifold $M$ to $\mathbb{R}^{2d+1}$,
where $t$ is time, is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of this modeling approach.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92}.  A
plot of the data from Figure~\ref{fig:ipc}, embedded following this
procedure, is shown in Figure~\ref{fig:embedding}.

 \begin{figure}
   \centering
\begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%   
  \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=\textwidth]{figs/colipc3d.png}
    \caption{\gcc}
    \label{fig:gccEmbedding}
  \end{subfigure}%

@@

%% can cut for space if need be:
The coordinates of each point on this plot are differently delayed
elements of the \col instructions per cycle time series
$X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
$X_{i+2\tau,obs}$ on the third, and so on.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding}---is an indication of
determinism%\footnote{A deeper analysis of
  %Figure~\ref{fig:embedding}---as alluded to on the previous
  %page---supports that diagnosis, confirming the presence of a chaotic
  %attractor in these cache-miss dynamics, with largest Lyapunov
  %exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
  %12-dimensional reconstruction space \cite{mytkowicz09}.}
  .  That structure can be used to build a forecast model.
 \subsubsection{LMA: Using dynamics in forecasting {\color{red} NOT EDITABLE}}

>>>>

\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA){\color{blue} EDITABLE}}\label{sec:lma}

When a generating process exhibits deterministic predictive nonlinear structure a common approach is to reconstruct the generating processes dynamics using a technique called delay-coordinate embedding and then use the geometric structure of this reconstruction as a proxy for the information transmission occurring in the underlying process to predict the future. In this section we discuss the theory and implementation of this forecasting scheme.

 \subsubsection{Reconstructing hidden dynamics {\color{blue} EDITABLE}}

Delay-coordinate embedding allows one to reconstruct a dynamical system's full
state-space dynamics from a \emph{single} scalar time-series
measurement---provided that some conditions hold regarding that data.
Specifically, if the underlying dynamics and the measurement
function---the mapping from the unknown state vector $\vec{X}$ to the
scalar value $X_{i,obs}$ that one is measuring---are both smooth and generic,
Takens~\cite{takens} formally proves that the delay-coordinate map
\[
F(\tau,m)(X) = ([X_{i+\tau,obs} ~ X_{i+\tau,obs} ~ \dots ~X_{i+m\tau,obs}])
\]
from a $d$-dimensional smooth compact manifold $M$ to $\mathbb{R}^{2d+1}$,
where $t$ is time, is a diffeomorphism on $M$---in other words, that
the reconstructed dynamics and the true (hidden) dynamics have the
same topology.

This is an extremely powerful result: among other things, it means
that one can build a formal model of the full system dynamics without
measuring (or even knowing) every one of its state variables.  This is
the foundation of this modeling approach.
The first step in the process is to estimate values for the two free
parameters in the delay-coordinate map: the delay $\tau$ and the
dimension $m$.  We follow standard procedures for this, choosing the
first minimum in the average mutual information as an estimate of
$\tau$ \cite{fraser-swinney} and using the false-near(est) neighbor
method of \cite{KBA92} to estimate $m$.  A
plot of the data from Figure~\ref{fig:ipc}, embedded following this
procedure, is shown in Figure~\ref{fig:embedding}.

 \begin{figure}
   \centering
\begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%   
  \begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/svd53dipc.png}
    \caption{\svdfive}
    \label{fig:svdfiveEmbedding}
  \end{subfigure}%   
  \begin{subfigure}{0.32\textwidth}
    \includegraphics[width=\textwidth]{figs/gcc3dipc.png}
    \caption{\gcc}
    \label{fig:gccEmbedding}
  \end{subfigure}%

@@

[[Update this paragraph if we include the second plot with \gcc]]
%% can cut for space if need be:
The coordinates of each point on this plot are differently delayed
elements of the \col instructions per cycle time series
$X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
$X_{i+2\tau,obs}$ on the third, and so on.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding}---is an indication of
determinism%\footnote{A deeper analysis of
  %Figure~\ref{fig:embedding}---as alluded to on the previous
  %page---supports that diagnosis, confirming the presence of a chaotic
  %attractor in these cache-miss dynamics, with largest Lyapunov
  %exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
  %12-dimensional reconstruction space \cite{mytkowicz09}.}
  .  That structure can be used to build a forecast model. 
 \subsubsection{LMA: Using dynamics in forecasting {\color{red} NOT EDITABLE}}

### 2014-03-05 pecomplexity.tex
finished the results section, mostly
moved a small chunk to concl
defined a newcommand
added usepackage{color}
adding sectoin on simple methods and cleaning
adding some figures and editing

\newcommand{\gcc}{{\tt 403.gcc}~}
\newcommand{\svd}{{\tt dgesdd}~}
\newcommand{\svdone}{{\tt dgesdd}$_1$~}
\newcommand{\svdtwo}{{\tt dgesdd$_2$}~}
\newcommand{\svdthree}{{\tt dgesdd$_3$}~}
\newcommand{\svdfour}{{\tt dgesdd$_4$}~}
\newcommand{\svdfive}{{\tt dgesdd$_5$}~}
\newcommand{\svdsix}{{\tt dgesdd$_6$}~}

@@

\begin{abstract}
[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
  Among other things, that implies that one should be able to use
  deterministic forecast rules to predict aspects of their behavior.
  That statement is sometimes---but not always---true. The memory and
  processor loads of some simple programs are easy to predict, for
  example, but those of more-complex programs like \gcc are not.
%%%%%%%%%%%%%%%%%%%%%%
%% I had to change all the \verb|blah| entries because they caused
%% latex to barf if they were in figure captions.  Odd bug.
%%%%%%%%%%%%%%%%%%%%%%
  The goal of this paper is to determine why that is the case. We
  conjecture that, in practice, complexity can effectively overwhelm
  the predictive power of deterministic forecast models. To explore
  that, we build models of a number of performance traces from
  different programs running on different Intel-based computers. We
  then calculate the \emph{permutation entropy}---a temporal entropy
  metric that uses ordinal analysis---of those traces and correlate
  those values against the prediction success.
\end{abstract}

>>>>

\newcommand{\gcc}{{\tt 403.gcc}~}
\newcommand{\svd}{{\tt dgesdd}~}
\newcommand{\svdone}{{\tt dgesdd}$_1$~}
\newcommand{\svdtwo}{{\tt dgesdd$_2$}~}
\newcommand{\svdthree}{{\tt dgesdd$_3$}~}
\newcommand{\svdfour}{{\tt dgesdd$_4$}~}
\newcommand{\svdfive}{{\tt dgesdd$_5$}~}
\newcommand{\svdsix}{{\tt dgesdd$_6$}~}
\newcommand{\naive}{na\"ive}

\parskip=3pt

@@

\begin{abstract}
[[Joshua: Needs to be rewritten to be more general and use computers as the application not the focus, should be done after we have a first draft of the paper]]  Computers are deterministic dynamical systems \cite{mytkowicz09}.
  Among other things, that implies that one should be able to use
  deterministic forecast rules to predict aspects of their behavior.
  That statement is sometimes---but not always---true. The memory and
  processor loads of some simple programs are easy to predict, for
  example, but those of more-complex programs like \gcc are not.
%%%%%%%%%%%%%%%%%%%%%%
%% I had to change all the \verb|blah| entries because they caused
%% latex to barf if they were in figure captions.  Odd bug.
%%%%%%%%%%%%%%%%%%%%%%
  The goal of this paper is to determine why that is the case. We
  conjecture that, in practice, complexity can effectively overwhelm
  the predictive power of deterministic forecast models. To explore
  that, we build models of a number of performance traces from
  different programs running on different Intel-based computers. We
  then calculate the \emph{permutation entropy}---a temporal entropy
  metric that uses ordinal analysis---of those traces and correlate
  those values against the prediction success.
\end{abstract}
